---
title: "Understanding Missing Data"
author: "Nick Tierney"
date: "11 July 2014"
output:
  html_document:
    highlight: haddock
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
  word_document: default
---

```{r global_options, include=FALSE}
library(knitr)
  # Set basic options. You usually do not want your code, messages, warnings etc.
  # to show in your actual manuscript however for the first run or two these will 
  # be set on.
opts_chunk$set(echo=TRUE, 
               warning=FALSE, 
               message=FALSE,
               cache = TRUE,
               include = TRUE,
               results = 'hide',
               dev = 'pdf')

## load necessary packages
library(rpart)
library(rpart.plot)
library(Amelia)
library(ggplot2)
library(ggdendro)
library(GGally)
library(grid)
library(mi)
library(xtable)

## add an if statement that installs the packages if they aren't there.
```

To run this code we need to make sure we have:

- The [supplementary material (brt.functions) found in Appendix S3](http://onlinelibrary.wiley.com/doi/10.1111/j.1365-2656.2008.01390.x/suppinfo) from the Elith et. al(2008) article.

For some reason in the code below I need to specify the full directory - even though this runs fine when I feed it into the console, it doesn't really work when I try to run a command like `simdat <- read.csv("sim.dat.csv")`.  So the code below specifies the rull directory.

```{r }
## load the MCAR.test function
  source('~/Dropbox/ALL THE THINGS/PhD/code/R/mex/R/mcar_test.R')

## source code from elith et al.
  source("~/Dropbox/ALL THE THINGS/PhD/Guides/Elith_BRT_Guide_2_of_2/brt.functions.R")

## this dataset has random missingness as well as missingness induced when the variable
## C1 is higher than 55.
sim.dat <- read.csv("~/Dropbox/ALL THE THINGS/PhD/code/R/mex/sim.dat.csv")

## depict the missingness.
missing.pattern.plot(sim.dat,
                     gray.scale = TRUE,
                     clustered = TRUE)
```

#### Data: 
A simulated missing dataset was created to explore the missing data methods.  This dataset `sim.dat.csv` is in the Github repo. This dataset contained five variables, two categorical and three continuous, with 1,000 observations in each. The two categorical factors, F1 and F2, ranged uniformly across categories nominally labelled 1-7, and 1-10, respectively.  The three continuous variables, named C1, C2 and C3, were normally distributed with means and standard deviations of 50 and 10, 90 and 10, and 30 and 3, respectively.

The code I wrote to create this data is below:

```{r data.create, results='markup'}

## make an id that goes from 1 thru to 1000, in increments of 1.
  ID <- seq(from = 1,
            to = 1000,
            by = 1)

# Now generate the variables.  There will be 3 continuous variables, and two factors.

## generate 3 missing continuous variables, C1..C3
  C1 <- rnorm(n = 1000, 
              mean = 50, 
              sd = 10)

#similar to FEV1% data
  C2 <- rnorm(n = 1000, 
              mean = 90, 
              sd = 10)

#similar to BMI data
  C3 <- rnorm(n = 1000, 
              mean = 30, 
              sd = 3)

## generate two factors.

## factor 1 - values between 1 and 7
## repeat 1...7, 1000 times)
  F1 <- rep(1:7, 
            length.out = 1000)

## make F1 a randomly ordered version of F1
  F1 <- sample(F1)

## factor 2 - values between 1 and 10
  F2 <- rep(x = 1:10, 
            times = 100)

## make F2 a randomly ordered version of F2
  F2 <- sample(F2)

## Collate variables: ID, C1...C3, F1, F2 into a dataframe, using data.frame
sim1 <- data.frame(ID, 
                   C1, 
                   C2, 
                   C3, 
                   F1, 
                   F2)
  
## double check the data
head(sim1)

dim(sim1)

class(sim1)

str(sim1)

```


I have written a function (`sim_miss_data.R`) to create the missing data but it needs to be changed so that I can save it to an object.

**R Datasets with missingness**

- airquality
- attenu

```{r}
missmap(airquality)

missmap(attenu)
```

### MCAR.test
Prior to identifying structure in the data, it is useful to ask whether there is sufficient missingness to warrant such an investigation - and to try and determine whether the data is missing completely at random (MCAR)

This can be done by splitting the data into two groups according to the presence or absence of a selected dependent variable, and to apply a t-test if the independent variables are continuous or a chi-square test if they are discrete, in order to determine equality of the means or the category probabilities, respectively. 

A Bonferroni adjustment or similar method can be used to allow for multiple tests.

I have written an R function `MCAR.test` which allows the user perform this test.  The function outputs a table giving the results of the t-test and the chi2 test. 

This function still currently not working, which is something I'll need to explore.

```{r mcar.test.chunk, eval=FALSE, include=FALSE, echo=TRUE}

sim.dat$F1 <- as.factor(sim.dat$F1)

sim.dat$F2 <- as.factor(sim.dat$F2)

sim.dat.2 <- sim.dat[,-c(1,7:9)]

head(sim.dat.2)

factor.list <- c("F1",
                 "F2")

MCAR.test(data = sim.dat,
          y = "C1",
          factor.list = factor.list)

```

### CART model.
We can then use CART models to predict the proportion of missing data in a row. For the CART model, I used the `rpart` and `rpart.plot` packages.

```{r cart.miss}

## this creates the rpart tree.
cart.small <- rpart(miss_perc ~ ID + C1 + C2 + C3 + F1 + F2,
                    data = sim.dat,
                    na.action = na.rpart, 
                    method = "anova")


## this actually plots the rpart tree.
prp(cart.small, 
    extra = 1, 
    type = 4, 
    prefix = "Prop. Miss = ")

```

### The BRT
For the BRT model, I use the `gbm` package, and the source code from elith et al.

```{r brt.example}

brt.mod <- gbm.step(data=sim.dat, 
                    gbm.x = c(3:7), 
                    gbm.y = 11,
                    family = "gaussian",
                    tree.complexity = 2,
                    learning.rate = 0.01,
                    bag.fraction = 0.5)

gbm.plot(brt.mod, 
         n.plots = 5, 
         write.title=F,
         plot.layout = c(2,3))

```

### Other existing Methods
Current solutions such `MissingDataGUI`, `VIM`, `missmap (in Amelia)`, `missing.pattern.plot (in mi)` usually focus on the user needing to visually search and find the trends. Whilst humans are very good at finding patterns, Having a model behind the output has more potential for really helping researchers explore their missing data problems in a precise way.  So whilst it is possible for people to use the methods provided in `mex`, it isn't necessarily easy, and straightforward.

`missmap` (from the `AMELIA II` package), which gives a visual depiction of the missingness in the dataset

```{r missmap}

missmap(sim.dat,
        rank.order = FALSE)
```

`missing.pattern.plot` is another missingness map tool that exists from the `im` package - this allows you to specify a "clustered" option, which groups data with similar missingness patterns together.

```{r im.map}

missing.pattern.plot(sim.dat,
                     gray.scale = TRUE)

```

`MissingDataGUI`

`VIM`

### Other Work?
Damjan has kindly pointed that the CART and BRT models may neglect useful information in the data's correlation structure.  An approach is to then apply hierarchical clustering to a dataset and then apply the CART or BRT on the dataset, predicting membership in a particular cluster.

So the code might look something like this.

#### Note:
This vignette needs to change the gbm.plot code- as this code- that I have modified in 'gbm.plot' - for the window() command - that I changed to X11() - it now plots everything after that in the X11 device, which I don't want, as it is slow.

```{r hclust.run}

## run the clustering on a dataset made of the 
hclust.fit <- hclust(dist(is.na.data.frame(sim.dat)))

require(ggdendro)

## plot the hclust
plot(hclust.fit)

## cut the hclust into 4 pieces.
c.id <- as.factor(cutree(hclust.fit, 4))

## add the 4 pieces into a dataframe
sim.clust <- data.frame(sim.dat, 
                        c.id)

names(sim.clust)

ggplot(data = sim.clust,
       aes(x = c.id,
           y = miss_perc)) + 
  geom_point()

head(sim.clust)

## get the CART model to predict membership in each cluster.
rpart.clust <- rpart(c.id ~ C1 + C2 + C3 + F1 + F2,
                    data = sim.clust,
                    na.action = na.rpart,
                    control = rpart.control(xval = 100))

prp(rpart.clust, 
    extra = 100, 
    type = 4, 
    prefix = "Clust ID = ")

```

The following code here makes a **nicer** looking dendro plot, but the code to do so is kinda complicated. 
```{r stack.ovflw.ggdendro}

data(mtcars)

x <- as.matrix(scale(mtcars))

dd.col <- as.dendrogram(hclust(dist(x)))

col.ord <- order.dendrogram(dd.col)

dd.row <- as.dendrogram(hclust(dist(t(x))))

row.ord <- order.dendrogram(dd.row)

xx <- scale(mtcars)[col.ord, row.ord]

xx_names <- attr(xx, "dimnames")

df <- as.data.frame(xx)

colnames(df) <- xx_names[[2]]

df$car <- xx_names[[1]]

df$car <- with(df, 
               factor(car, 
                      levels = car, 
                      ordered=TRUE))

mdf <- melt(df, 
            id.vars="car")

## extract dendrogram data and create plots

ddata_x <- dendro_data(dd.row)

ddata_y <- dendro_data(dd.col)

### Set up a blank theme
theme_none <- theme(
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  panel.background = element_blank(),
  axis.title.x = element_text(colour=NA),
  axis.title.y = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y = element_blank(),
  axis.line = element_blank()
  #axis.ticks.length = element_blank()
)

### Create plot components ###    
# Heatmap
p1 <- ggplot(mdf, 
             aes(x = variable, 
                 y = car)) + 
  geom_tile(aes(fill = value)) +
  scale_fill_gradient2()

# Dendrogram 1
p2 <- ggplot(segment(ddata_x)) + 
  geom_segment(aes(x = x, 
                   y = y, 
                   xend = xend, 
                   yend = yend)) + 
  theme_none + 
  theme(axis.title.x = element_blank())

# Dendrogram 2
p3 <- ggplot(segment(ddata_y)) + 
  geom_segment(aes(x = x, 
                   y = y, 
                   xend = xend, 
                   yend = yend)) + 
  coord_flip() +
  theme_none

## Use grid graphics and some manual alignment to position the three plots on the page

### Draw graphic ###

grid.newpage()
print(p1, vp=viewport(0.8, 
                      0.8, 
                      x = 0.4, 
                      y = 0.4))

print(p2, 
      vp = viewport(0.52, 
                    0.2, 
                    x = 0.45, 
                    y = 0.9))

print(p3, 
      vp = viewport(0.2, 
                    0.8, 
                    x = 0.9, 
                    y = 0.4))

```

## My Hopes?
That we can bundle all of these functions together into a package that generates some nifty plots and diagnostics for the user.  But what I think would make this really nifty would be to create plots in ggplot format and even better, to generate plots in ggvis format, so that other users can explore their missing data in an interactive way.   I think that a really nice goal that could come out of our work, which would have to be in the later stages, once everything is working, would be to make it easier for researchers to produce publication-quality visualisations from using our package.  For example, making a nicer dendrogram plot, or to make a nicer


